---
title: "Day 1 notes"
output: html_document
---

This document contains the code written in class on 09/19/2026.

First, load the libraries 

```{r}
library(tidyverse)
library(lme4)
library(emmeans)
```

## Demonstrating the analysis of an RCBD

1. Load the data 
```{r}
url <- "https://raw.githubusercontent.com/k-state-id3a/mixed-models-fall26/refs/heads/main/data/cochrancox_kfert.csv"
df <- read.csv(url)

df$rep <- as.factor(df$rep)
df$K2O_lbac <- as.factor(df$K2O_lbac)
```

```{r}
df |> 
  ggplot(aes(K2O_lbac, yield))+
  geom_point()+
  theme_classic()
```

2.1. Fit an all-fixed model.
```{r}
m_fixed <- lm(yield ~ K2O_lbac + rep, data = df)
summary(m_fixed)
```

1.2. Fit a mixed-effects model. 
```{r}
m_random <- lmer(yield ~ K2O_lbac + (1|rep), data = df)
summary(m_random)
```

```{r}
sigma(m_fixed)
sigma(m_random)
```

3. Check the marginal means for the K2O treatments, and note that the means do not differ between models 1 and 2, but the SE and CIs do. 

```{r}
(mg_means_fixed <- emmeans(m_fixed, ~K2O_lbac, contr = list(c(1, 0, 0, 0, -1))))
(mg_means_random <- emmeans(m_random, ~K2O_lbac, contr = list(c(1, 0, 0, 0, -1))))
```

## Demonstrating the analysis of an RCBD - II

1. Load the data 

```{r}
df_rcbd <- agridat::gomez.seedrate
df_rcbd$rate <- as.factor(df_rcbd$rate)
str(df_rcbd)
```

2. Fit three models: 

- No blocks (we used this one to demonstrate how blocks increase the precision in the results)
- Blocks as fixed 
- Blocks as random 

```{r}
m1_noblocks <- lm(yield ~ 1 + rate, data = df_rcbd)

m1_fixed <- lm(yield ~ 1 + rate + rep, data = df_rcbd)
emmeans(m1_fixed, ~rep)

m1_random <- lmer(yield ~ 1 + rate + (1|rep), data = df_rcbd)
getME(m1_random, "b")
```

Compare the $\sigma^2_\varepsilon$
```{r}
sigma(m1_noblocks)
sigma(m1_fixed)
VarCorr(m1_random)
```


3. Compare the means 

```{r}
emmeans(m1_noblocks, ~rate)
```

```{r}
emmeans(m1_fixed, ~rate)
```

```{r}
emmeans(m1_random, ~rate)
```



## Demonstrating the analysis of a split-plot design

(split-plot in an RCBD)

1. Load the data 

```{r}
dat <- agridat::yates.oats
dat$nf <- factor(dat$nitro)

str(dat)
```

2. Fit the model. Remember this is a split-plot in an RCBD with gen in the whole plot and nitrogen in the subplot. 
```{r}
m2 <- lmer(yield ~ 1 + gen + nf + gen:nf +
             (1|block) + (1|block:gen) ,
           data = dat)

hist(resid(m2))
plot(predict(m2), resid(m2))
```

3.1. Check out how, when we use mixed models, the degrees of freedom are "automatically" adjusted. 

```{r}
car::Anova(m2, test.statistic = "F")
```

3.2. Check out DF and SE for the factors in the whole plot and in the split plot. Remember that there are more true reps for the factor in the split-plot.  

```{r}
emmeans(m2, ~gen, contr = list(c(1, -1, 0)))
emmeans(m2, ~nf, contr = list(c(1, -1, 0, 0)))
```

## Demonstrating the analysis of a repeated measures design

1. Load the data 
```{r}
options(contrasts = c("contr.treatment", "contr.poly"))
url <- "https://raw.githubusercontent.com/k-state-id3a/mixed-models-fall26/refs/heads/main/data/temperature_2.csv"
library(glmmTMB)
dd_temp <- read.csv(url)
dd_temp$Pig <- as.factor(dd_temp$Pig)
dd_temp$Treatment <- as.factor(dd_temp$Treatment)
dd_temp$Pen <- as.factor(dd_temp$Pen)
dd_temp$Time <- as.factor(dd_temp$Time)
dd_temp$Temperature_F <- (dd_temp$Temperature_C *9/5) + 32
```

1.1. How many pigs per EU? Just one (n column). 
```{r}
dd_temp |> 
  group_by(Treatment, Pen, Time) |> 
  summarise(n())
```

2.1. Fit the model with a Compound symmetry covariance structure. 
```{r}
m3 <- lmer(Temperature_F ~ Treatment * Time + (1|Pen),
           data = dd_temp |> arrange(Pen, Time))
summary(m3)


Z <-  getME(m3, "Z") 
sigma2_s <- as.data.frame(VarCorr(m3))[1,4]
G <- diag(c(rep(sigma2_s, length(getME(m3, "b") ))))
```


```{r}
Z %*% G %*% t(Z) %>%
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column("x") %>% 
  pivot_longer(cols = `1`:`350`, names_to = "y") %>%
  mutate(x = as.numeric(x), y= as.numeric(y)) %>% 
  
  left_join(
  diag(sigma(m3)^2, nrow = nrow(dd_temp)) %>% 
  as.data.frame() %>% 
  rownames_to_column("x") %>% 
  pivot_longer(cols = V1:V350, names_to = "y") %>%
  mutate(y = str_replace(y, "V", "") %>% as.numeric(),
         x = as.numeric(x)) %>% 
  rename(s2 = value)) %>% 
  filter(abs(x)<51, abs(y)<51) |> 

  ggplot(aes(-x,y))+
  geom_tile(aes(fill = value+s2))+
  theme_minimal()+
  scale_fill_viridis_c()+
  theme(aspect.ratio = 1, 
        axis.text = element_blank(), 
        legend.position = "none", 
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.ticks = element_blank())+
  labs(title = "zoom-in for a variance-covariance\nmatrix for a CS")

```


3. Try other covariance structures. 
```{r}
m_repeated <- glmmTMB(Temperature_C ~ Treatment * Time + cs(0 + Time |Pen),
              family = gaussian(link = "identity"),
              REML = TRUE,
              data = dd_temp)

summary(m_repeated)


m_repeated <- glmmTMB(Temperature_C ~ Treatment * Time + ar1(0 + Time |Pen),
              family = gaussian(link = "identity"),
              REML = TRUE,
              data = dd_temp)
DHARMa::simulateResiduals(m_repeated, plot = T)

m_repeated <- glmmTMB(Temperature_C ~ Treatment * Time + us(0 + Time |Pen),
              family = gaussian(link = "identity"),
              REML = TRUE,
              data = dd_temp)
DHARMa::simulateResiduals(m_repeated, plot = T)



m_repeated <- glmmTMB(Temperature_F ~ Treatment * Time + us(0 + Time |Pen),
              family = gaussian(link = "identity"),
              REML = TRUE,
              data = dd_temp)
DHARMa::simulateResiduals(m_repeated, plot = T)
```

```{r}
m_repeated
```




